In this section, I will be creating some maps of the data. I'll start by loading some packages and the dataframes.

```{r}
library(here)
library(tidyverse)
library(RColorBrewer)

#loading a spatial object for the contiguous US. 

states <- map_data("state")

df_fit <- readRDS(here("data/processed_data/df_fit_alt.rds"))
```

I do a bit more cleaning to the data to tidy things up and create some new variables for the best fitting model. Spatial methods were very marginally better than nonspatial methods.
```{r}
df_fit <- df_fit %>%
  mutate(region = tolower(State)) %>% 
 # replace(is.na("Ratio Observed/ Expected"), 0) %>% 
  #ungroup() %>% 
  mutate(dykes_chi = ((`Death Cases` - `Expected Death Cases`)/ sqrt(`Expected Death Cases`)),
         ## GLM
         predicted_observed_mult = mult_fit/`MRR`,
         predicted_observed_mult = ifelse(!is.finite(predicted_observed_mult), 0, predicted_observed_mult),
         observed_predicted_mult = `MRR`/mult_fit,
         observed_predicted_mult = ifelse(!is.finite(observed_predicted_mult), 0, observed_predicted_mult),
         ## CAR
         predicted_observed_car = car_fit/`MRR`,
         predicted_observed_car = ifelse(!is.finite(predicted_observed_car), 0, predicted_observed_car),
         observed_predicted_car = `MRR`/car_fit,
         observed_predicted_car = ifelse(!is.finite(observed_predicted_car), 0, observed_predicted_car))
```

Next, I will merge my dataframe contain the variables of interest with the spatial object. 
```{r}
df_sf_fit <- merge(states,df_fit, by = "region")
# This next step is import to ensure that no "holes" appear in the map.
df_sf_fit <- df_sf_fit[order(df_sf_fit$order), ]
```

Now I'm ready to start mapping! I'll begin by looking at the map of the raw HIV death rate across to contiguous US by selected race/ ethnicity.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Death Rate (estimated)`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "HIV-Related Death Rates
Among Non-Hispanic Black and White Persons
United States, 2020", x = "", y = "", fill = "Death rate per 100,000") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/raw_rate_map.png"), units="in", width=10, height=4, dpi=300)
```

The map for HIV-related death differs drastically across the United States by race/ethnicity. In most states, the highest HIV-related death were among Black person. This is especially true in the South region. Non-coastal western states consistently had lowest rates HIV-related deaths across the selected racial/ethnic groups, expect among multi-racial non-Hispanic persons. Non-Hispanic White persons were the only of the selected racial/ ethnic group that did not have a HIV-related death rate higher than ~10 per 100,000.

```{r}
mrr_raw <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`MRR`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "a. Raw", x = "", y = "", fill = "Ratio") +
  scale_fill_distiller(palette = "RdYlGn", limits = c(0,16)) +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/mrr_map.png"), units="in", width=10, height=4, dpi=300)
```

```{r}
map_1 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living in Poverty`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living in Poverty", x = "", y = "", fill = "Percent")  +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_2 <-ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Median Household Income`/1e4)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Median Household Income", x = "", y = "", fill = "Per $10,000")  +
  theme(plot.title= element_text(hjust = 0.5))
```

```{r}
map_3 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Uninsured Under 65 Years Old`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Uninsured Under 65 Years Old", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```


```{r}
map_4 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Without an Internet subscription`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Without an Internet Subscription", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_5 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percent Living with Food Insecurity`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percent Living with Food Insecurity", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
map_6 <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill =`Percentage Diagnosed Late`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)   + theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "Percentage Diagnosed Late", x = "", y = "", fill = "Percent") +
  theme(plot.title= element_text(hjust = 0.5)) 
```

```{r}
library(gridExtra)
library(ggpubr)
grid_map <- grid.arrange(map_2, map_1, map_3, map_4, map_5, map_6, ncol = 2, nrow = 3)
grid_map <- annotate_figure(grid_map, top = text_grob("Maps of Selected Predictors, United States, 2020", 
               color = "black", face = "bold", size = 14))


ggsave(here("results/grid_map.png"), grid_map, units="in", width=11, height=10, dpi=300)
```
`

Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = MRR_null_loc)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  scale_fill_distiller(palette = "RdYlGn", limits = c(0,16)) + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Expected Black-to-White MRR with Null Effect of Location, 
       United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
```
Next, let's map a probability map based on where we'd expect deaths if risk of HIV-related death was consistent across the US and race.
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = `Null Expected Rate`/`Null Expected Rate`)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  scale_fill_distiller(palette = "RdYlGn") + theme_void() + 
  theme(strip.text = element_text(size = 14),
    strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Expected Black-to-White MRR with Null Effect of Location and Race, United States, 2020", x = "", y = "", fill = "Ratio") +
  theme(plot.title= element_text(hjust = 0.5))
```

Next, let's using Dykes and Unwin's method of smooth map using a chi square value. 
```{r}
ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = dykes_chi)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) +
  facet_wrap(. ~ Race)  + theme_void() + 
  theme(strip.text = element_text(size = 14),
        strip.background = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        axis.text = element_blank(),
        legend.position = "bottom") +
  labs(title = "Dykes and Unwin's Chi Square by Race/ Ethnicity, United States, 2020", x = "", y = "", fill = "Chi Square") +
  scale_fill_distiller(palette = "RdYlGn") +
  theme(plot.title= element_text(hjust = 0.5))

ggsave(here("results/chi_map.png"), units="in", width=10, height=4, dpi=300)
```
##Dykes and Unwin’s chi-square statistic
Dykes and Unwin (Dykes and Unwin 2001) technique  has the effect of creating a distribution of values closer to normal (as opposed to a Poisson distribution of rates and counts encountered thus far). We can therefore apply a diverging color scheme where green hues represent less than expected rates and red hues represent greater than expected rates.

```{r}
##Smoothed map using fitted values from best model (CAR)
car_smooth <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = car_fit)) +
   scale_fill_distiller(palette = "RdYlGn", limits = c(0,16)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5)  + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "b. Smooth (fitted)", x = "Ratio", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/car_smooth.png"))
``` 

Next, let's look at the map for the ratio of predicted to observed (P/O) values. 
```{r}
car_pred <- ggplot(df_sf_fit, aes(long, lat)) +
  geom_polygon(aes(group = group, fill = predicted_observed_car)) +
  scale_fill_distiller(palette = "RdYlGn", limits = c(0,3)) +
  coord_map("albers",  lat0 = 45.5, lat1 = 29.5) + 
  theme_void() + 
  theme(strip.text = element_text(size = 14),
        legend.position = "bottom") +
  labs(title = "c. Predicted/ Observed", x = "", y = "", fill = "") +
  theme(plot.title= element_text(hjust = 0.5))
ggsave(here("results/car_pred.png"))
``` 

```{r}
grid_mrr <- grid.arrange(mrr_raw, car_smooth, car_pred, ncol = 3, nrow = 1)

grid_mrr <- annotate_figure(grid_mrr, top = text_grob("Maps of Black-to-White HIV-Related Mortality Rate Ratio, United States, 2020", 
               color = "black", face = "bold", size = 14))
ggsave(here("results/grid_mrr.png"), grid_mrr, units="in", width=10, height=4, dpi=300)
```
We can see that the model fitted pretty well. A higher P/O ratio (more red values) indicates that the model overestimated the observed values, while a lower P/O ratio (more green values) indicates that the model underestimates the observed values.

