```{r message=FALSE}
library(here)
library(GGally)
library(skimr)
library(tidyverse)
```

```{r}
df <- readRDS(here("data/processed_data/dataframe.rds"))
df_sdoh20 <- readRDS(here("data/processed_data/df20.rds"))
```

Let's collapse age groups to just look at 20-44 due to issue with very small populations among some groups. let also look at hiv deaths over entire time period.
```{r}
df_sub <- df %>% 
  filter(Age_Group %in% c("20-29", "30-44"))

df_sub <- df_sub %>%
  group_by(District, Race_Ethnicity, Sex) %>% 
  summarise(HIV_Deaths = sum(HIV_Deaths),
            Population = sum(Population),
            Uninsured = mean(`% Uninsured`),
            Poverty_Rate = mean(Poverty_Rate_1620),
            n_PCP = sum(`# Primary Care Physicians`),
            mean_PCP_rate = mean(`PCP Rate`)) 
```

Next, let's find the expected number of HIV deaths. First, we'll calculate lambda.
```{r}
lambda <- sum(df_sub$HIV_Deaths)/sum(df_sub$Population)

df_sub <- df_sub %>%
  mutate(Expected_Deaths = lambda * Population)
```

Then, we'll calculate rates for observed HIV deaths and expected HIV deaths.
```{r}
df_sub <-  df_sub %>% 
  mutate(HIV_Rate = ((HIV_Deaths + 0.5)/Population) *1e5)

df_sub <-  df_sub %>% 
  mutate(Expected_Rate = ((Expected_Deaths + 0.5)/Population) *1e5)
```

Next, let's find the standardized mortality ratio.
```{r}
df_sub <- df_sub %>%
  mutate(SMR = HIV_Rate/Expected_Rate)
```

Next, I looked at the distribution of SMR
```{r}
df_sub %>% 
  ggplot() +
  geom_histogram(aes(x = SMR), bins = 10)

summary(df_sub$SMR)
```
The histogram revealed a very right-skewed distribution.

Let's take a look at HIV rate and SMR more closely by running some model diagnostics in ggfortify.
```{r}
model <- lm(df_sub$HIV_Rate ~ 1)
model_SMR <- lm(df_sub$SMR ~ 1)

library(ggfortify)

autoplot(model)
autoplot(model_SMR)
```


Let's see if any transformations can normalize this distribution. I decided to start with log(x +1) since the distribution was so skewed.
```{r}
df_sub %>% 
  mutate(logSMR = log(SMR + 1)) %>% 
  ggplot() +
  geom_histogram(aes(x = logSMR), bins = 20)
```
But based on visual assessment, it appears that this approach was not enough to normalized the distribution! After some quick Googling, it appears some potential options are using a mixture model or Box-Cox transformation. 

I'll try the Box-Cox transformation first. chrome-extension://efaidnbmnnnibpcajpcglclefindmkaj/https://cran.r-project.org/web/packages/spmoran/vignettes/vignette_spmoran_nongaussian.pdf
```{r}
library(MASS)
b <-  boxcox(lm(df_sub$SMR ~ 1))

lambda2 <- b$x[which.max(b$y)]

df_sub <- df_sub %>%
  ungroup() %>%
  mutate(exact_x = (df_sub$SMR^lambda2 - 1) / lambda2)

exact_model <- lm(df_sub$exact_x ~ 1)
plot(exact_model)

df_sub %>% 
  ggplot() +
  geom_histogram(aes(x = exact_x), bins = 20)

shapiro.test(df_sub$exact_x)
```
With the p = 0.25, we do not find statistically significant evidence to reject normality. Great!

Zero-inflated model. We can try this on the count data..
```{r}
library(pscl)

m1 <- zeroinfl(HIV_Deaths ~ Uninsured | 1,  offset = Population,  data = df_sub, dist = "poisson")
m2 <- zeroinfl(HIV_Deaths ~ . | 1,  offset = Population,  data = df_sub, dist = "negbin")

summary(m1)
summary(m2)
```
Admittedly, I am bit unsure of what route to take next. I am considering 
I am considering pivoting to HIV prevalence data???
https://bmcinfectdis.biomedcentral.com/articles/10.1186/1471-2334-14-285


```{r}
corr_cols <- 
ggpairs(corr_cols)

```



Let's look at HIV-related death rates by race/ethnicity.
```{r message=FALSE}
df_sub %>% 
  ggplot() +
  geom_histogram(aes(x = exact_x), bins = 20) + facet_wrap (. ~ Race_Ethnicity) 

df_sub %>%
  group_by(Race_Ethnicity) %>%
  skim(exact_x)
```

Then by sex...
```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (. ~ Sex) 

df_sub %>%
  group_by(Sex) %>%
  skim(HIV_Rate)
```

And by race/ ethnicity by sex.
```{r message=FALSE, warning=FALSE}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (Sex ~ Race_Ethnicity) 

df_sub %>%
  group_by(Race_Ethnicity, Sex) %>%
  skim(HIV_Rate)
```


Next, let's get a sense of how things look by District. 
```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (. ~ District) 

df_sub %>%
  group_by(District) %>%
  skim(HIV_Rate)
```

```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
geom_histogram(aes(x = HIV_Rate), bins = 20) + facet_wrap (District ~ Sex) 

df_sub %>%
  group_by(District) %>%
  skim(HIV_Rate)
```


```{r}
df_sub <- df_sub %>% mutate(logHIV_Rate = log10(HIV_Rate + 1))
```

```{r}
df_sub %>% 
  drop_na(HIV_Rate) %>% 
  ggplot() +
  geom_histogram(aes(x = HIV_Rate), bins = 20)


df_sub %>% 
  drop_na(HIV_Rate) %>%
  ggplot() +
  geom_histogram(aes(x = logHIV_Rate))
```

```{r}
x_test <- df %>%
  drop_na(HIV_Rate) %>%
  filter(HIV_Rate > 0)

shapiro.test(x_test$HIV_Rate)
```

```{r}

```

